---
title: "US Food and Drug Administration-approved drugs"
author: "Mohammed Ali, Ali Ezzat"
date: "December 31, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## FDA-approved drugs target
- Get FDA approved drugs.

I can verify the groups from here (https://www.drugbank.ca/drugs), I will need to update the approved number to see how it is done
```{r fda}
drugs <- read_csv("data/drugs.csv")
drugs_groups <- read_csv("data/drug_groups.csv")
drugs$fda_approved <- is.na(drugs$fda_label)
table(drugs_groups$text)
barplot(table(drugs_groups$text))
```


- Create figure 1 sections

## A Drug–Target Network
- Get FDA-approved drugs and their targets.
- Generate a bipartite graph of drug–protein interactions
 In which a drug and a protein are connected to each other if the protein is a known target of the
drug
- Generate two biologically relevant network projections. 
 In the *drug network*, nodes represent drugs, and two drugs are connected to each other if they share at least one target protein.
 In the complementary *target protein network* (TP network), nodes are proteins, and two proteins
are connected if they are both targeted by at least one common drug.

## Definitions
### Bipartite Graph
A bipartite graph, also called a bigraph, is a set of graph vertices decomposed into two disjoint sets such that no two graph vertices within the same set are adjacent [mathworld](http://mathworld.wolfram.com/BipartiteGraph.html). 

### Anatomical Therapeutic Chemical (ATC) 
In the Anatomical Therapeutic Chemical (ATC) classification system, the active substances are divided into different groups according to the organ or system on which they act and their therapeutic, pharmacological and chemical properties. 
Drugs are classified in groups at five different levels. [who](https://www.who.int/medicines/regulation/medicines-safety/toolkit_atc/en/)

### Human Interactome
The human interactome is the set of protein–protein interactions (the interactome) that occur in human cells [Wikipedia](https://en.wikipedia.org/wiki/Human_interactome).