---
title: "Journey to the Center of DrugBank XML"
author: "Mohammed Ali, Ali Ezzat"
date: "December 31, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.tree)
```

## Objective
We will try to take a journey together to explore the wonders of *DrugBank* xml database. How the drugs information structured in different ways inside it and how we can retrive these information using **R**. So our purpose is to show how to parse the database and not to be overwhelmed with the domain information. Let us begin.

## What is DrugBank database
The DrugBank database is a comprehensive, freely accessible, online database containing information on drugs and drug targets.As both a bioinformatics and a cheminformatics resource, DrugBank combines detailed drug (i.e. chemical, pharmacological and pharmaceutical) data with comprehensive drug target (i.e. sequence, structure, and pathway) information [Wikipedia](https://en.wikipedia.org/wiki/DrugBank) and it is in *XML* format.


## Do you remember your children names
Each drug in the database is represented by `drug` node which contains beside its associated attributes many children which differ in name, functionality and more importantly to us *structure*. Let us take a peak:

```{r drug_sructure}
drug <- Node$new("Drug")
drug$AddChild("Drugbank Id(s)")
drug$AddChild("Name")
drug$AddChild("description")
drug$AddChild("Groups")
drug$AddChild("References")
drug$AddChild("synonyms")
drug$AddChild("products")
drug$AddChild("mixtures")
drug$AddChild("categories")
drug$AddChild("Many others ...")
plot(drug)
```

Wow that is a lot