---
title: "Drug Bank Browser"
author: "Mohammed Ali; Ali Ezzat"
date: "February 4, 2019"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(collapsibleTree)
require(colorspace)
library(DT)
library(data.tree)
library(tidyverse)
library(lubridate)
library(forcats)
library(collapsibleTree)

## Read drugs
drugs <- read_csv(file = "data/drugs.csv")
drugs$type <- as.factor(drugs$type)
drugs$state <- fct_explicit_na(drugs$state, "Unknown")
## Read Groups
drug_groups <- read_csv("data/drug_groups.csv")
drugs_all <- drugs %>%
  select(name, primary_key, type , state, created) %>%
  full_join(drug_groups, by = c("primary_key" = "parent_key")) %>%
  rename(group = "text") %>%
  mutate(created_year = year(created),
         created_month = month(created)) %>%
  group_by(created_year, created_month, group, state)

drugs_df <- reactiveValues(data = drugs_all %>%
                             select(name, type, state, group, created_year, created_month)) 

```

## Objective
In this tutorial we will build a shiny application to browse [Drug Bank Database](https://www.drugbank.ca) drugs main features based on our developed package [dbparser](https://dainanahan.github.io/dbparser/index.html), [Collapsible Tree package](https://github.com/AdeelK93/collapsibleTree) and [Shiny DataTable package](https://rstudio.github.io/DT/).

Through this tutorial you will learn:

- How to parse Drugbank database and save desired dataset from it.
- Visualize data using *Collapsible Tree* using dynamic features.
- Build Dynamic filters to display tree selected node rows.


## Parse DrugBank database
Dealing with `dbparser` is quite straightforward as all what you need to do is:

- downloading the public `DrugBank` database,
- then, giving its physical location to `get_xml_db_rows` method.
- and findally, calling your favourit parser to return a regular csv drug related dataset.

```{r eval=FALSE, highlight=TRUE}
library(dbparser)
library(reader)
## parse data from XML and save it to memory
get_xml_db_rows("drugbank_record.xml")

## load drugs data
drugs <- parse_drug()

## load drug groups data
drug_groups <- parse_drug_groups()

## save datasets to use them later
write_csv(drugs, "data/drugs.csv")
write_csv(drugs, "data/drug_groups.csv")
```

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r eruptions, echo=FALSE}
     inputPanel(
      selectInput(
        "hierarchy", "Tree hierarchy",
        choices = c(
          "created_year", "created_month", "group", "state"
        ),
        selected = c("group", "state"),
        multiple = TRUE
      )
    )

renderCollapsibleTree({
    drugs_all %>%
      summarize(`Number of Drugs` = n()) %>%
      collapsibleTreeSummary(
        hierarchy = input$hierarchy,
        root = "Drugs",
        inputId = "node",
        width = 800,
        attribute = "Number of Drugs"
      )
})
  
  observeEvent(input$node, {
    tree_list <- input$node
    fltr_str <- ""
    if (!is.null(tree_list[["created_year"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "created_year", "=='", tree_list[["created_year"]], "'")
    }
    
    if (!is.null(tree_list[["created_month"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str,
               "created_month",
               "=='",
               trimws(tree_list[["created_month"]]),
               "'")
    }
    
    if (!is.null(tree_list[["group"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "group", "=='", tree_list[["group"]], "'")
    }
    
    if (!is.null(tree_list[["state"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "state", "=='", tree_list[["state"]], "'")
    }
    if (fltr_str == "") {
      drugs_df$data <- drugs_all
    } else {
      drugs_df$data <- drugs_all %>% filter_(fltr_str)
    }
    
  })
  
DT::renderDataTable({
    datatable(drugs_df$data, options = list(pageLength = 5))
})
  
  
  
  

add_and <- function(x) {
  if (x != "") {
    x <- paste0(x, "&")
  }
  return(x)
}

construct_filter <- function(key, value, prev_filter = "") {
  if (!is.null(value)) {
    prev_filter <- add_comma(prev_filter)
    prev_filter <- paste0(prev_filter, key, "='", value, "'")
  }
}
```



