---
title: "Browsing the DrugBank Database"
author: "Mohammed Ali; Ali Ezzat"
date: "February 4, 2019"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(collapsibleTree)
require(colorspace)
library(DT)
library(data.tree)
library(tidyverse)
library(lubridate)
library(forcats)
library(collapsibleTree)

## read drugs
drugs <- read_csv(file = "data/drugs.csv")

## convert drug type to factors, we will use that later
drugs$type <- as.factor(drugs$type)

## if the drug state is NA, then mark it as "Unknown"
drugs$state <- fct_explicit_na(drugs$state, "Unknown")

## read groups
drug_groups <- read_csv("data/drug_groups.csv")
```

## Objective
In this tutorial, we will build a Shiny application to browse the *[DrugBank](https://www.drugbank.ca)* database. Specifically, we will display drugs-related information that is stored in *DrugBank* with the help of the 
following packages: main features based on our developed package

- *[dbparser](https://dainanahan.github.io/dbparser/index.html)*, a package that 
    we developed earlier for parsing the XML-formatted *DrugBank* database
- *[collapsibleTree](https://github.com/AdeelK93/collapsibleTree)*, an 
    [htmlwidget](http://www.htmlwidgets.org/) that displays Reingold-Tilford 
    tree diagrams using D3.js
- *[DataTables (DT)](https://rstudio.github.io/DT/)*, which takes R data frames 
    and displays them as tables in HTML pages

Through this tutorial you will learn:

- How to parse the *DrugBank* database and retrieve the desired data from it.
- Visualize data using `collapsibleTree` using dynamic features.
- Build dynamic filters to display table rows corresponding to the selected 
    tree node.

--------------

## Parsing the *DrugBank* database
Using `dbparser` is quite straightforward. What we need to do is:

- Download the *DrugBank* database. Latest release is freely available for 
    download [here](https://www.drugbank.ca/releases/latest). At the time of 
    writing this post, the version of *DrugBank* is "**5.1.2**".
- Pass the physical location of the downloaded data to the 
    `dbparser::get_xml_db_rows()` method.
- Call the desired parser function to return an R data frame containing the 
    intended drugs-related information.
- Save retrieved data in CSV format.

```{r eval=FALSE, highlight=TRUE}
library(dbparser)
library(readr)

## parse data from DrugBank's XML and save it to memory
get_xml_db_rows("drugbank_database.xml")

## retrieve drugs basic data
drugs <- parse_drug()

## retrieve drugs groups data
drug_groups <- parse_drug_groups()

## save datasets to use them later
write_csv(drugs, "data/drugs.csv")
write_csv(drug_groups, "data/drug_groups.csv")
```

--------------

## Visualize drugs data with `collapsibleTree`
To *browse* our drugs data, we first load the data.

```{r eval=FALSE, highlight=TRUE}
## load drugs basic data
drugs <- read_csv(file = "data/drugs.csv")

## convert drug 'type' to factors (we will use that later)
drugs$type <- as.factor(drugs$type)

## if the drug 'state' is NA, then mark it as "Unknown"
drugs$state <- fct_explicit_na(drugs$state, "Unknown")

## load drugs groups data
drug_groups <- read_csv("data/drug_groups.csv")
```

We have `r nrow(drugs)` different drugs within `r nrow(drug_groups)` different 
groups. Browsing all these data in a regular data table can be a little 
inconvenient. Hence, we shall build a *visual* means of filtering the data to 
enable users to focus only on the subset of drugs that they want. That is where 
the `collapsibleTree` package comes in.

Now, we will build the tree's data frame with the features that we wish to 
filter our main data frame with:

- `group`: filter drugs based on their *groups* {`r unique(drug_groups$text)`}
- `state`: filter drugs based on their *states* {`r levels(drugs$state)`}
- `created_year`:  filter drugs based on which *year* they were inserted into the database
- `created_month`: filter drugs based on which *month* they were inserted into the database

```{r}
drugs_all <- drugs %>%
  ## The features we want to use on our filters and display in the table
  select(name, primary_key, type , state, created) %>%

  ## join datasets using drug primary key
  full_join(drug_groups, by = c("primary_key" = "parent_key")) %>%

  ## name properly the group feature
  rename(group = "text") %>%

  ## extract the year and month of drug database insertion
  mutate(created_year = year(created),
         created_month = month(created))
```

We then display the tree levels so that the user can arrange them dynamically 
to filter the drugs based on her/his preferences.

```{r}
inputPanel(
  ## input so user can build the tree filters based on her/his preference
  selectInput (
    "hierarchy",
    "Tree hierarchy",
    choices = c("group", "state", "created_year", "created_month"),
    selected = c("group", "state"),    ## initially selected tree levels
    multiple = TRUE)
)
```

The final step is to build the tree itself based on the new dataset.

```{r}
renderCollapsibleTree({
  drugs_all %>%

  ## specify (for collapsibleTree) which fields to filter in the displayed tree with
  group_by(created_year, created_month, group, state) %>% 
  summarize() %>%

  ## display tree
  collapsibleTreeSummary(
    hierarchy = input$hierarchy,
    root = "Drugs",
    inputId = "node",               ## will be used later to get filter values
    width = 800,
    tooltip = FALSE
  )
})
```

--------------

## Display Results

At this point, we want the user to be able to filter the data using the 
displayed tree. Specifically, whenever the user clicks a node, we want the 
retrieved drugs to be filtered by the clicked node. 

For example, if the tree is showing only two levels, `state` followed by 
`created_year`, then clicking on the *solid* state would get all the *solid* 
drugs in the *DrugBank* database. However, if the year *2016* under the *solid* 
state was clicked instead, then only solid drugs created in *2016* would be 
returned.

```{r echo=FALSE}
drugs_df <- reactiveValues(data = drugs_all %>%
                             ## the filter features
                             select(name, type, state, group, created_year, created_month)) 

  ## tree clicked node
  observeEvent(input$node, {
    tree_list <- input$node
    ## filter statement that we are building
    fltr_str <- ""
    if (!is.null(tree_list[["created_year"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "created_year", "=='", tree_list[["created_year"]], "'")
    }
    
    if (!is.null(tree_list[["created_month"]])) {
      ## join multiple filters
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str,
               "created_month",
               "=='",
               trimws(tree_list[["created_month"]]),
               "'")
    }
    
    if (!is.null(tree_list[["group"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "group", "=='", tree_list[["group"]], "'")
    }
    
    if (!is.null(tree_list[["state"]])) {
      fltr_str <- add_and(fltr_str)
      fltr_str <-
        paste0(fltr_str, "state", "=='", tree_list[["state"]], "'")
    }
    if (fltr_str == "") {
      drugs_df$data <- drugs_all
    } else {
      drugs_df$data <- drugs_all %>% filter_(fltr_str)
    }
    
  })

DT::renderDataTable({
    datatable(drugs_df$data, options = list(pageLength = 5))
})

add_and <- function(x) {
  if (x != "") {
    x <- paste0(x, "&")
  }
  return(x)
}

```

So, first we build the *reactive* data frame for the data table. 
It has to be reactive as it is changeable via the clicked tree nodes used for
filtering. 

```{r eval=FALSE}
drugs_df <- reactiveValues(data = drugs_all %>%
                               ## the filter features
                                select(name, type, state, group, created_year, created_month)) 
```

Now, we observe the tree node `click` event to apply filtering to the drugs 
data table.

```{r eval=FALSE}
## tree clicked node
observeEvent(input$node, {
    tree_list <- input$node
    
    ## filter statement that we are building
    fltr_str <- ""
    if (!is.null(tree_list[["created_year"]])) {
        fltr_str <- add_and(fltr_str)
        fltr_str <-
        paste0(fltr_str, "created_year", "=='", tree_list[["created_year"]], "'")
    }
    
    if (!is.null(tree_list[["created_month"]])) {
        ## join multiple filters
        fltr_str <- add_and(fltr_str)
        fltr_str <-
            paste0(fltr_str,
                   "created_month",
                   "=='",
                   trimws(tree_list[["created_month"]]),
                   "'")
    }
    
    if (!is.null(tree_list[["group"]])) {
        fltr_str <- add_and(fltr_str)
        fltr_str <- paste0(fltr_str, "group", "=='", tree_list[["group"]], "'")
    }
    
    if (!is.null(tree_list[["state"]])) {
        fltr_str <- add_and(fltr_str)
        fltr_str <-paste0(fltr_str, "state", "=='", tree_list[["state"]], "'")
    }
    if (fltr_str == "") {
        drugs_df$data <- drugs_all
    } else {
        drugs_df$data <- drugs_all %>% filter_(fltr_str)
    }
})
```

This is the small helper function that we used to join multiple conditions.

```{r eval=FALSE}
add_and <- function(x) {
    if (x != "") {
        x <- paste0(x, "&")
    }
    return(x)
}
```

Finally, we display the filtered data table.

```{r eval=FALSE}
DT::renderDataTable({
    datatable(drugs_df$data, options = list(pageLength = 5))
})
```

-----

## Conclusion
In this post, we showed how to use the `dbparser` package to build an 
interactive browser for the *DrugBank* database. This post is also a 
prelude to our upcoming application, *DrugBank Browser*, that is intended to 
be used for all sorts of neat exploratory analysis stuff, so stay tuned!

-----

## About the Authors
We are a team of data scientists called *[Dainanahan](http://dainanahan.github.io)*. 
Our aim is to provide practitioners in the field of drug discovery with 
useful tools that would make their life easier and help increase their 
productivity. This suite of tools that we are planning to make is codenamed 
*[DrugVerse](https://www.researchgate.net/project/DrugVerse-A-Suite-of-Tools-for-Use-in-Drug-Discovery)*.